---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# veganomics

<!-- badges: start -->
<!-- badges: end -->

The goal of veganomics is to provide extra functions for using vegan with -omics data such as RNA-Seq or microarrays, including functions for:

- Filtering, normalizing and scaling expression matrices for use with rda()
- Improved plotting of variance partioning between constrained/non-constrained components from rda()
- Improved plotting of study design and sample strain from rda()

Some functions also work for cca()

## Installation

You can install the development version of veganomics from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("MalteThodberg/veganomics")
```

## Examples using Bioconductor datasets

The following includes some example analysis using various datasets from Bioconductor.

```{r package}
library(veganomics)
theme_set(theme_bw())
```

### RDA on bladderbatch data

```{r bladderbatch}
# Data
library(SummarizedExperiment)
library(bladderbatch)
data("bladderdata")

# Convert from eSet to SummarizedExperiment and format a bit
SE <- as(bladderEset, "SummarizedExperiment")
SE$batch <- factor(SE$batch) |> forcats::fct_infreq()
SE$cancer <- relevel(SE$cancer, "Normal")

# Normalize
EM <- t(assay(SE))

# Simple PCA
fit_pca <- rda(EM, scale=TRUE)
plotTotal(fit_pca, color=SE$cancer, shape=SE$batch)

# Fit RDA
fit_rda <- rda(formula=EM~outcome,
							 data=as.data.frame(colData(SE)), scale=TRUE)

# Visualize decomposition
plotDecomposition(fit_rda, 
									color=SE$outcome, 
									shape=SE$batch,
									color_scale = scale_color_brewer("Cancer", palette = "Set1"),
									shape_scale = scale_shape("Batch"))

# Further inspect model
plotDesign(fit_rda, color=SE$cancer, shape=SE$batch) +
	scale_color_brewer("Cancer", palette = "Set1") +
	scale_shape("Batch")

plotStrain(fit_rda, color=SE$cancer, shape=SE$batch) +
	scale_color_brewer("Cancer", palette = "Set1") +
	scale_shape("Batch")

# Example of corresponding tests in vegan
anova(fit_rda, 
			permutations = how(nperm=99), # SET THIS TOO ATLEAST 999!!!
			by="term")

anova(fit_rda, 
			permutations = how(nperm=99), # SET THIS TOO ATLEAST 999!!!
			by="axis")
```

### pRDA on parathyroidSE data

```{r parathyroidSE}
library(parathyroidSE)
data("parathyroidGenesSE")

# Filter counts using edgeR
CM <- filterCounts(~treatment*time,
						 data=colData(parathyroidGenesSE),
						 y=assay(parathyroidGenesSE))

# Normalize
EM <- normalizeCounts(CM)

# Simple PCA
fit_pca <- rda(EM, scale=TRUE)
plotTotal(fit_pca, color=parathyroidGenesSE$patient)

# RDA
fit_rda <- rda(formula=EM~treatment*time,
								data=as.data.frame(colData(parathyroidGenesSE)), scale=TRUE)

# Visualize decomposition
plotDecomposition(fit_rda, color=parathyroidGenesSE$treatment, shape=parathyroidGenesSE$time,
									color_scale = scale_color_brewer("Treatment", palette = "Set1"),
									shape_scale = scale_shape("Time"))

# Highligh batch effect
plotDecomposition(fit_rda, color=parathyroidGenesSE$patient,
									color_scale = scale_color_brewer("Patient", palette = "Set2"))

# Fit partial RDA to remove batch effect
fit_prda <- rda(formula=EM~treatment+time+treatment:time+Condition(patient),
		data=as.data.frame(colData(parathyroidGenesSE)), scale=TRUE)

# Corrected
plotDecomposition(fit_prda, color=parathyroidGenesSE$treatment, shape=parathyroidGenesSE$time,
									color_scale = scale_color_brewer("Treatment", palette = "Set1"),
									shape_scale = scale_shape("Time"))
plotDecomposition(fit_prda, color=parathyroidGenesSE$patient,
									color_scale = scale_color_brewer("Patient", palette = "Set2"))

# Further introspection of pRDA
plotDesign(fit_prda, color=parathyroidGenesSE$treatment, shape=parathyroidGenesSE$time) +
	scale_color_brewer("Treatment", palette = "Set1") +
	scale_shape("Time")

plotStrain(fit_prda, color=parathyroidGenesSE$treatment, shape=parathyroidGenesSE$time) +
 	scale_color_brewer("Treatment", palette = "Set1") +
 	scale_shape("Time")

# Test significance of terms
anova(fit_prda, 
			permutations = how(nperm=99), # SET THIS TOO ATLEAST 999!!!
			by="term")
```
